---
title: "Inferência Bayesiana 1"
date: "2026-10-09"
lang: "pt-BR"
author: Diogo Melo
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
      mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js"
---

```{r setup, echo=FALSE}
library(rmarkdown)
knitr::opts_chunk$set(eval = FALSE, fig.asp = 1)
# clipboard
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "Copiar código <i class=\"fa fa-clipboard\"></i>",
    success_text = "Copiado! <i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "Não copiado <i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

# Defining a very simple posterior

Consider the following model:

Likelihood:

$$
y_i \sim Normal(\mu, \sigma = 1) 
$$

Prior:

$$
\begin{aligned}
\mu &\sim Normal(0, 1) \\
\sigma &\sim Exponential(1)
\end{aligned}
$$

For very simple models, like this one, we can write and evaluate the posterior directly.
We can then use these calculated values to plot the full posterior.

Lets write this posterior in R:

In the following code, we will just walk through all the possible values of $\mu$ and measure their posterior probability using the product of the prior by the likelihood (or sum in the log scale). This will give us the shape of the posterior distribution, which we can then plot and inspect to draw conclusions.

```{r}
set.seed(2)
# Data:
y = rnorm(100, mean = 2, sd = 1)

# Posterior, in log scale for numerical stability
logP_theta_y = function(mu, sigma = 1, data = y){
  ll = sum(dnorm(data, mu, sigma, log = TRUE))
  logprior = dnorm(mu, 0, 1, log = TRUE) + dexp(sigma, 1, log = TRUE)
  logprior + ll # P(theta)P(y|theta)
}
```

In this univariate case it's easy to check many possible values of
$\mu$ to find the high probability values, but in high dimensions this
is not feasible.

```{r}
x = seq(0, 4, 0.01)
px = Vectorize(logP_theta_y)(x)
plot(px ~ x, pch = 19, col = 2, xlab = "mean parameter (mu)", ylab = "proportional to P(mu|y)")
abline(v = 2)
```

Given this posterior, what is the most probable value of the parameter?

# Installing the Rethinking package

These instructions are reproduced from: [https://github.com/rmcelreath/rethinking](https://github.com/rmcelreath/rethinking). See this website for more details and other options.

There are three steps. (1) Install the C++ toolchain, (2) install ``cmdstanr``, (3) install ``rethinking``. Details follow.

First, install the C++ toolchain. Go to ``https://mc-stan.org/docs/cmdstan-guide/cmdstan-installation.html#cpp-toolchain`` and follow the instructions for your platform.

Second, install the ``cmdstanr`` package. Visit ``https://mc-stan.org/cmdstanr/`` for details. 
To install ``cmdstanr``, run:
```
# we recommend running this in a fresh R session or restarting your current session
install.packages("cmdstanr", repos = c('https://stan-dev.r-universe.dev', getOption("repos")))
```

The first time you install cmdstanr, you will also need compile the libraries with ``cmdstanr::install_cmdstan()``. All this of this bother is worth it. You just have to do it once. 

Third, you can install ``rethinking`` from within R using:
```
install.packages(c("coda","mvtnorm","remotes","loo","dagitty","shape","rstanarm"))
remotes::install_github("rmcelreath/rethinking")
```

Note that the ``rethinking`` package is not on CRAN, just on github. The ``rethinking`` package is never going to be on CRAN. So if you get an error about rethinking not being available for your version of R, it is because you tried to install from CRAN. Use the github code above.

# Defining a model in the rethinking package

The rethinking syntax closely mirrors the mathematical notation we have been using in class.
For example, a simple linear regression of $y_i$ on $x_i$ using some standard-ish priors can be defined as:

$$
\begin{aligned} 
  y_i &\sim Normal(\mu_i, \sigma) \\
  \mu_i &= \alpha + \beta x_i  \\
  \alpha &\sim Normal(0, 1) \\
  \beta &\sim Normal(0, 1) \\
  \sigma &\sim Exponential(1) \\
\end{aligned} 
$$
First, let's do the exercise of visualizing the assumed prior distributions for the three parameters.

A normal distribution with $\mu = 0$ and $\sigma = 1$:

```{r}
curve(dnorm(x, mean = 0, sd = 1), -3, 3)
```

An exponential distribution with $\lambda = 1$

```{r}
curve(dexp(x, rate = 1), 0, 3)
```

What are these prior distributions telling us? What is the range of possible values for these two parameters?

Next, lets we simulate some data:

```{r }
N = 100 # sample size

# Linear regression parameters
alpha = 1 # Intercept
beta = 1  # slope on x
## Simulate a normal response that has mu as  a linear function of a predictor 
x = rnorm(N)           # Simulated independent variable. Mean = 0 and sd = 1
mu = alpha + beta * x  # linear model for the expected value
y = rnorm(N, mu)       # Response variable, with residual sd = 1

```

And now we fit our model, that translates to the following code:

```{r, eval = FALSE, message=FALSE, warning=FALSE}

library(rethinking)

## Rethinking code
m1 = ulam(alist(
  y ~ normal(mu, sigma),     # Likelihood
  mu <- a + b * x,           # Linear model
  a ~ normal(0, 1),          # Prior on intercept
  b ~ normal(0, 1),          # Prior on slope
  sigma ~ exponential(1)),   # Prior on residual variance
  data = list(y = y, x = x), # Data list
  iter = 2000, chains = 4, cores = 4, refresh = 0, cmdstan = TRUE) # Sampler control parameters

```

The `precis()` function generates a simple summary table of fitted coefficients:

```{r }
precis(m1, digits = 2) # Summary of model fit
```

__Exercise:__ Fit the same model using the lm() function in R, and
compare the summary of the two fitted models.

If the sample size is large, the likelihood overwhelms the prior and
even strong priors can lead to similar inferences.

__Exercise:__ Change the priors. Can you find a prior that
substantially changes the estimates? Next, reduce the sample size
(change the N variable). Do the priors make a difference now?


# Fitting models using Bayesian methods

In the previous section, we fitted a linear model using the OLS method.
Now, we are going to fit the same model using a Bayesian approach.

The Bayesian approach to linear regression is based on the posterior
distribution of the parameters $\alpha$ and $\beta$. This posterior
distribution is proportional to the product of the likelihood of the
data given the parameters and the prior distribution of the parameters.

The likelihood of the data given the parameters is the same as the one
used in the OLS method. The prior distribution of the parameters is 
usually a normal distribution with mean zero and a fixed standard deviation.

We will generate a new data set with known parameters to illustrate some differences between the OLS and Bayesian methods.

```{r create_dataset2}
set.seed(4) 

x = runif(50, 50, 100)
y = 1000 + 150 * x + rnorm(50, 0, 1000)
data = data.frame(x, y)
```

Always visualize your data before fitting a model

```{r plot2}
plot(y ~ x, data = data)
```

first, lets fit the model using the OLS method:

```{r fit_ols}
mod_ols <- lm(y ~ x, data = data)

# The summary function provides a lot of information about the model, 
# including estimates, standard errors, p-values for the coefficients, 
# and residual standard deviation (sigma)
summary(mod_ols)
```

Now, let's fit the model using the rstanarm package, which fits Bayesian models using the Stan language and provides a nice set of standard weekly informative priors.

```{r fit_bayesian, messages=FALSE, eval=FALSE}
suppressPackageStartupMessages(library(rstanarm))
mod_bayesian <- stan_glm(y ~ x, data = data)

summary(mod_bayesian)
```

Let's compare the coefficient estimates of the two methods.

```{r compare, eval=FALSE}
coef(mod_ols)
coef(mod_bayesian)

library(ggplot2)

# Extract intercept and slope for both models
ols_intercept <- coef(mod_ols)[1]
ols_slope <- coef(mod_ols)[2]
bayesian_intercept <- coef(mod_bayesian)[1]
bayesian_slope <- coef(mod_bayesian)[2]


ggplot(data, aes(x, y)) + 
  geom_point() +
  geom_abline(intercept = ols_intercept, slope = ols_slope, color = "red", linetype = "solid") +
  geom_abline(intercept = bayesian_intercept, slope = bayesian_slope, color = "blue", linetype = "dashed") + 
  theme_bw()

```

How similar are the estimates of the two methods?

### Explicit model in the rethinking package

Now, let's use the rethinking package to fit the model using some standard priors we used in class.

```{r fit_rethinking, message=FALSE, eval=FALSE}
suppressPackageStartupMessages(library(rethinking))

mod_rethinking = ulam(
  alist(
    y ~ normal(mu, sigma),
    mu <- a + b * x,
    a ~ normal(0, 1),
    b ~ normal(0, 1),
    sigma ~ exponential(1)
  ),
  data = data, chains = 4, cores = 4
)
precis(mod_rethinking)
```

How are the estimates now? Awful!! 

What is happening is that the default priors in the rethinking code are too far away from the true values of the parameters. One way to avoid this type of problem is to scale the variables before fitting the model.

```{r scale, eval=FALSE}
data$x_scaled = scale(data$x)
data$y_scaled = scale(data$y)

mod_rethinking_scaled = ulam(
  alist(
    y_scaled ~ normal(mu, sigma),
    mu <- a + b * x_scaled,
    a ~ normal(0, 1),
    b ~ normal(0, 1),
    sigma ~ exponential(1)
  ),
  data = data, chains = 4, cores = 4
)
mod_stanarm_scaled = stan_glm(y_scaled ~ x_scaled, data = data)
mod_ols_scaled = lm(y_scaled ~ x_scaled, data = data)
```

Now, let's compare the estimates of the two models.

```{r compare_scaled, eval=FALSE}
coef(mod_ols_scaled)
coef(mod_stanarm_scaled)
coef(mod_rethinking_scaled)
```

Scalling variables is a good practice when fitting models with Bayesian methods. It improves computational efficiency and make defining reasonable priors easier.

# Extending linear regression

We can extend the simple linear regression model to accommodate multiple predictors. These are the **multiple linear regression** models, of which the simple linear regression is just a particular case with a single predictor variable.

In the next sections, we will show how to fit multiple regression models using the `rethinking` package, interpret the model coefficients, and visualize the results.

# Additive multiple linear model

In an additive multiple linear regression, the expected value of the response $μ$ is the sum of multiple predictor variables $X_i$, each one weighted by its own coefficient $β_i$ :

$$
\begin{align}
y_i &\sim Normal(\mu_i, \sigma) \\
\mu_i & = \alpha + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_j x_{ij} \\
\end{align}
$$

## Fitting

We will use the dataset of the biennial plant *Ipomopsis aggregata*. Forty plants were assigned to two treatments: unprotected from grazing and protected from grazing. The fruit yield of each plant was measured along with the diameter of the rootstock.

Prepare the data:

```{r data, eval=TRUE}
df <- read.csv("https://raw.githubusercontent.com/lmmx/crawley-r-statistics-book-notes/master/code/ipomopsis.csv")
names(df) <- c("Root", "Fruit", "Grazing")
df$Root <- scale(df$Root, center = TRUE, scale = TRUE)
df$Fruit <- scale(df$Fruit, center = TRUE, scale = TRUE)

# Ungrazed is the control, and fenced grazed is the fenced treatment:
df$Grazing0 <- ifelse(df$Grazing == "Ungrazed", 1, 0)
```

Fit the additive model using `rethinking`:

```{r fit_model}
# Fitting the model using quap
m1 <- ulam(
  alist(
    Fruit ~ dnorm(mu, sigma),
    mu <- a + b * Grazing0 + c * Root,
    a ~ normal(0, 1),  
    b ~ normal(0, 1),   
    c ~ dnorm(0, 1),  
    sigma ~ exponential(1)          
  ),
  data = df, chains = 4, cores = 4
)
```

Extract and summarize the coefficients:

```{r summarize_model}
precis(m1)

m1_coef <- precis(m1)[1]
```

## Visualizing Predictions

Create a plot to show the observed data and predicted values:

```{r plot_predictions}
# Plot the observed data
plot(df$Fruit ~ df$Root, 
     col = ifelse(df$Grazing0 == 1, "red", "black"), 
     pch = ifelse(df$Grazing0 == 1, 17, 16),
     xlab = "Basal diameter (z-scores)", 
     ylab = "Fruit dry weight (z-scores)")

# Add the predicted lines
abline(a = m1_coef["a", ], b = m1_coef["b", ], col = "black", lwd = 2)
abline(a = m1_coef["a", ] + m1_coef["c", ], b = m1_coef["b", ], col = "red", lwd = 2)

# Add legend
legend("topleft", legend = c("Fenced", "Control"), col = c("red", "black"), 
       pch = c(17, 16), lty = 1, bty = "n")

```

# Adding Interactions to the Model

Interactions can model how the relationship between predictors and the response variable changes depending on the levels of other predictors. For example, we can model an interaction between soil moisture (`water`) and shade (`shade`) in the `tulips` dataset.

Prepare the data:

```{r tulips_data}
data(tulips)
tulips$blooms <- tulips$blooms / max(tulips$blooms)
tulips$water <- scale(tulips$water, scale = FALSE)
tulips$shade <- scale(tulips$shade, scale = FALSE)
```

Fit the interaction model:

```{r fit_interaction}
m2 <- quap(
  alist(
    blooms ~ dnorm(mu, sigma),
    mu <- b0 + b1 * water + b2 * shade + b3 * water * shade,
    b0 ~ dnorm(0, 1),
    b1 ~ dnorm(0, 1),
    b2 ~ dnorm(0, 1),
    b3 ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = tulips
)
```

Summarize the interaction model:

```{r summarize_interaction}
precis(m2)
```

Visualize the interaction:

```{r visualize_interaction}
post <- extract.samples(m2)
plot(blooms ~ water, col = 2:4, pch = 16, data = tulips)
for (s in unique(tulips$shade)) {
  abline(
    a = mean(post$b0) + mean(post$b2) * s,
    b = mean(post$b1) + mean(post$b3) * s,
    col = s + 1
  )
}
legend("topright", legend = c("Shade 1", "Shade 2", "Shade 3"), col = 2:4, pch = 16)
```

# Robust regression

One key advantage of using Bayesian methods is that it is very easy to modify our models using the exact same syntax we used in the simple models. This process can be taken up to absurd extremes, leading to fastatically complicted models that are tailor-made for your specific problems.

Let's illustrate this point here by defining robust regression. One common problem in linear regression models is outliers. Sometimes these points are measuremente errors, but sometimes they are just really true manifestations of biological processes that generate variation that is outside the expected normal (gaussian) range of the population we are studing, and removing them would be a mistake. 

To illustrate, let's use a dataset in which a single point can shift the estimates of the parameters of our simple linear regression.

## A high-leverage observation in real data

The `trees` dataset in R contains measurements of girth, height, and
volume for 31 black cherry trees. We will model volume as a linear
function of girth. This is a deliberately simple model: the largest
tree may have a real biological explanation, and its influence may also
signal that a straight line is an inadequate description of the
relationship.

First, use an ordinary linear model as a diagnostic to locate the case
with the greatest leverage. Leverage is determined by how unusual a
predictor value is; it is not the same as a large residual. The common
rule $2p/n$ is only a heuristic for flagging cases, not a deletion rule.

```{r trees_diagnostic, eval=TRUE}
data(trees, package = "datasets")
diagnostic_fit <- lm(Volume ~ Girth, data = trees)
case_high_leverage <- which.max(hatvalues(diagnostic_fit))
leverage_reference <- 2 * length(coef(diagnostic_fit)) / nrow(trees)

data.frame(
  case = case_high_leverage,
  girth = trees$Girth[case_high_leverage],
  volume = trees$Volume[case_high_leverage],
  leverage = hatvalues(diagnostic_fit)[case_high_leverage],
  reference_2p_over_n = leverage_reference,
  cooks_distance = cooks.distance(diagnostic_fit)[case_high_leverage]
)
```

Case 31 has the largest girth and is flagged for high leverage. Its
Cook's distance also indicates that it can influence the fitted line.
We will keep the case in the data for the first fit, then omit it only
as a sensitivity analysis.

## Fitting both models with `quap`

Fit the same Gaussian regression to the full dataset and to the dataset
without case 31. Both models use the same weakly informative priors.

```{r trees_quap_fits, eval=TRUE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(rethinking))

trees_without_case <- trees[-case_high_leverage, ]

m_trees_all <- quap(
  alist(
    Volume ~ dnorm(mu, sigma),
    mu <- a + b * Girth,
    a ~ dnorm(0, 100),
    b ~ dnorm(0, 10),
    sigma ~ dexp(0.1)
  ),
  data = trees
)

m_trees_without_case <- quap(
  alist(
    Volume ~ dnorm(mu, sigma),
    mu <- a + b * Girth,
    a ~ dnorm(0, 100),
    b ~ dnorm(0, 10),
    sigma ~ dexp(0.1)
  ),
  data = trees_without_case
)

precis(m_trees_all, digits = 2)
precis(m_trees_without_case, digits = 2)
```

Compare the posterior estimates of the slope `b` and residual standard
deviation `sigma`. How much do they change when case 31 is omitted? The
comparison is a sensitivity analysis; it does not establish that the
case should be discarded.

## Visualizing the influence on the fit

The lines below show posterior-mean fits from both models on the same
scatterplot. The highlighted case is present in the full-data fit and
excluded from the second fit.

```{r trees_fit_comparison, eval=TRUE, fig.asp=1}
posterior_all <- extract.samples(m_trees_all)
posterior_without_case <- extract.samples(m_trees_without_case)
girth_values <- seq(min(trees$Girth), max(trees$Girth), length.out = 100)
point_colors <- ifelse(seq_len(nrow(trees)) == case_high_leverage,
                       "firebrick", "grey35")

plot(Volume ~ Girth, data = trees, pch = 19, col = point_colors,
     xlab = "Girth (inches)", ylab = "Volume (cubic feet)")
text(trees$Girth[case_high_leverage], trees$Volume[case_high_leverage],
     labels = paste("Case", case_high_leverage), pos = 2, col = "firebrick")
lines(girth_values,
      mean(posterior_all$a) + mean(posterior_all$b) * girth_values,
      col = "navy", lwd = 2)
lines(girth_values,
      mean(posterior_without_case$a) +
        mean(posterior_without_case$b) * girth_values,
      col = "darkorange3", lwd = 2, lty = 2)
legend("topleft",
       legend = c("All trees", "Without case 31", "High-leverage case"),
       col = c("navy", "darkorange3", "firebrick"),
       lty = c(1, 2, NA), pch = c(NA, NA, 19), bty = "n")
```

Does the slope change in the direction you expected from the position of
case 31? Notice that a high-leverage observation can strongly affect the
fit even when it is a genuine measurement. Before removing such a point,
investigate its source and consider whether the model's functional form
is adequate. The plotted lines are posterior means; they do not show the
uncertainty around either fit.

# Modifying the model to be less sensitive to outliers

One advantage of Bayesian modeling is that we can change the residual
distribution while keeping the same linear predictor. A model with
heavier-tailed residuals can reduce the influence of observations that
are unusually far from the fitted line. More flexible models can also be
unnecessarily complex, so the added flexibility should address a
specific problem.

Some outliers are measurement errors; others are genuine observations
from biological processes with more extreme variation than a Gaussian
model allows. Removing a genuine observation can discard important
information. Instead, we will keep case 31 and replace the Gaussian
residual distribution with a Student-$t$ distribution. We will then
compare the sensitivity of each model by fitting it both with and
without that case.

## Student-t residual model

The Student-$t$ distribution has heavier tails than the Normal
distribution. Its degrees of freedom, $\nu$, control the tail weight:
smaller values allow more probability for observations far from the
model's expected value. As $\nu$ increases, the distribution approaches
the Normal. In the model below, $\nu = 2 + \exp(\text{log\_nu})$ keeps
the degrees of freedom above 2, so the residual variance is finite.

```{r trees_student_fits, eval=TRUE, message=FALSE, warning=FALSE}
m_trees_student_all <- quap(
  alist(
    Volume ~ dstudent(nu, mu, sigma),
    mu <- a + b * Girth,
    nu <- 2 + exp(log_nu),
    a ~ dnorm(0, 100),
    b ~ dnorm(0, 10),
    sigma ~ dexp(0.1),
    log_nu ~ dnorm(1, 1)
  ),
  data = trees
)

m_trees_student_without_case <- quap(
  alist(
    Volume ~ dstudent(nu, mu, sigma),
    mu <- a + b * Girth,
    nu <- 2 + exp(log_nu),
    a ~ dnorm(0, 100),
    b ~ dnorm(0, 10),
    sigma ~ dexp(0.1),
    log_nu ~ dnorm(1, 1)
  ),
  data = trees_without_case
)

precis(m_trees_student_all, digits = 2)
precis(m_trees_student_without_case, digits = 2)
```

## Comparing sensitivity to case 31

Compare the change in the posterior mean slope when case 31 is omitted
under each residual distribution. The fits below use posterior draws
from the same four models shown in the previous sections.

```{r trees_sensitivity, eval=TRUE}
set.seed(42)
post_normal_all <- extract.samples(m_trees_all)
post_normal_without <- extract.samples(m_trees_without_case)
post_student_all <- extract.samples(m_trees_student_all)
post_student_without <- extract.samples(m_trees_student_without_case)

slope_change_normal <- abs(mean(post_normal_all$b) - mean(post_normal_without$b))
slope_change_student <- abs(mean(post_student_all$b) - mean(post_student_without$b))

data.frame(
  residual_distribution = c("Normal", "Student-t"),
  slope_all_cases = c(mean(post_normal_all$b), mean(post_student_all$b)),
  slope_without_case_31 = c(mean(post_normal_without$b), mean(post_student_without$b)),
  absolute_slope_change = c(slope_change_normal, slope_change_student)
)
```

Is the change in slope smaller under the Student-$t$ model? In this
example it is smaller, but not zero: heavy-tailed residuals reduce the
influence of the unusual response, while the extreme predictor value
can still give case 31 leverage.

## Plotting the four fitted lines

The solid lines use all 31 trees; the dashed lines omit case 31. Compare
the separation between each pair of lines to the slope changes in the
table above.

```{r trees_student_comparison, eval=TRUE, fig.asp=1}
girth_values <- seq(min(trees$Girth), max(trees$Girth), length.out = 100)
point_colors <- ifelse(seq_len(nrow(trees)) == case_high_leverage,
                       "firebrick", "grey35")

plot(Volume ~ Girth, data = trees, pch = 19, col = point_colors,
     xlab = "Girth (inches)", ylab = "Volume (cubic feet)")
text(trees$Girth[case_high_leverage], trees$Volume[case_high_leverage],
     labels = paste("Case", case_high_leverage), pos = 2, col = "firebrick")

lines(girth_values,
      mean(post_normal_all$a) + mean(post_normal_all$b) * girth_values,
      col = "navy", lwd = 2)
lines(girth_values,
      mean(post_normal_without$a) + mean(post_normal_without$b) * girth_values,
      col = "navy", lwd = 2, lty = 2)
lines(girth_values,
      mean(post_student_all$a) + mean(post_student_all$b) * girth_values,
      col = "firebrick", lwd = 2)
lines(girth_values,
      mean(post_student_without$a) + mean(post_student_without$b) * girth_values,
      col = "firebrick", lwd = 2, lty = 2)

legend("topleft",
       legend = c("Normal: all cases", "Normal: without case 31",
                  "Student-t: all cases", "Student-t: without case 31",
                  "High-leverage case"),
       col = c("navy", "navy", "firebrick", "firebrick", "grey35"),
       lty = c(1, 2, 1, 2, NA), pch = c(NA, NA, NA, NA, 19), bty = "n")
```

The Student-$t$ model is less sensitive here if its two fitted lines are
closer together than the two Normal-model lines. This does not mean that
Student-$t$ residuals make high-leverage cases harmless: they mainly
reduce the effect of large residuals. Always investigate an influential
case and check whether the model's functional form is appropriate.